---
title: "Project: Forecasting Gas prices"
name: "Daniel Freeland"
date: "May 13, 2021"
output:  github_document
---
```{r}
library(pacman)
p_load(fpp3, tidyverse, readr, tidyr)
```

**Finding Data**
The Energy Information Administration (EIA) is responsible for the US Department of Energy’s statistics and data. There is a wealth of data on crude oil and gasoline on the EIA’s website but it can be tricky to access it. To help with this, the EIA created an API for ease of access. I utilized the API at [eia website](www.eia.com/opendata). The EIA releases new data weekly, monthly, or annually depending on the different types of data. I obtained code from the following website to download the JSON file and saved the data as `eia_monthly_data.txt`.

[code for pulling data](https://www.analyticsvidhya.com/blog/2021/01/use-r-to-pull-energy-data-from-the-department-of-energys-eia-api/)

```{r, message = FALSE}
# The Raw Data
eia_months <- read_csv("eia_months.csv")
eia_months <- eia_months %>% mutate(Date = yearmonth(Date)) 
tsibble(eia_months) %>% head(n=5)
tsibble(eia_months) %>% tail(n=5)
```

To make graphing easier I renamed the columns 

```{r}
eia_months <- tsibble(eia_months, index = "Date") %>% 
  rename(Unleaded = "Unleaded Regular Gasoline, U.S. City Average Retail Price, Monthly",
         Crude = "Imported  Crude Oil Real Price, Monthly",
         PDI = "Real Disposable Personal Income, Monthly",
         Unemp = "Civilian Unemployment Rate, Monthly" ,
         CPI = "Consumer Price Index (all urban consumers), Monthly") 
```

## 1 Regression
To do a forecast using regression I regressed The US average unleaded gasoline price on the one month lag Crude oil price to forecast 12 months (from April 2020 to March 2021). I had to add more regressors and after exploring the correlation plots and testing many models for the highest adjusted R-squared and lowest AICc I selected `Crude_PDI_Season` as the best model which regresses gas price on lag 1 crude oil price and lag 1 disposable personal unemployment with a seasonal component of 1 year.  The residuals looked good and the forecasts captured the movement of the time series for the new data. 

Here is the plot for the past 10 years of average price of unleaded gasoline in US from the Energy Information Administration (EIA) . 

```{r}
eia_months %>% autoplot(Unleaded) + geom_line(color = "red") + 
  labs(x="", y = "Unleaded",
       title = "TIME PLOT OF U.S. GASOLINE PRICES, IN DOLLARS PER GALLON",
        subtitle = "(Jauary 2010 - March 2021)")
```

The average price of unleaded gas time series looks like a random walk. The series shows a sharp drop in the level around Jan 2015, but does not appear to have a strong seasonal pattern. The ACF shows a scalloped shape as the correlations decay. 

```{r}
eia_months %>% ACF(Unleaded, lag_max= 36) %>%
  autoplot()
```
Look at a seasonal plot shows striking similarities for the same month over 10 years. The blue line shows that the average price is higher in the Spring/Summer months and lower in the Fall/Winter months.
```{r}
eia_months %>% gg_subseries(Unleaded)
```

I chose to look at the lagged 1 month regressor of Crude price. I used `mutate()` to include the lagged regressors by 1 month assuming a forecast would use the previous months data which have a strong positive correlation of 0.951. The prices have different ranges so I displayed them separately. 

```{r, warning = FALSE, message = FALSE}
eia_months_lag <- eia_months %>% 
  mutate(L1_Crude = lag(Crude, n = 1),
         L1_Unemp = lag(Unemp, n = 1),
         L1_PDI   = lag(PDI, n = 1),
         L1_CPI   = lag(CPI, n = 1))

eia_months_lag %>%  pivot_longer(c(Unleaded, L1_Crude), names_to ="Series") %>% 
  ggplot(aes(Date, value)) + geom_line() +
  facet_grid(vars(Series), scales = "free_y")
```

```{r, warning = FALSE, message=FALSE}
eia_months_lag %>% 
 GGally::ggpairs(columns = c(2, 8)) + 
  labs(title = "GASOLINE PRICES AGAINST LAGGED PRICE OF CRUDE OIL")
```



I fit a regression with`TSLM` using `L1_crude`, a superimposed time series plots is given to show that it seems a good candidate for.  The first row contains missing values for the lagged variables so I did not include it. The training set goes up to March 2020. The testing set is from April 2020 to March 2021. 


```{r}
fit_Unleaded_1 <- eia_months_lag %>% select(Date,Unleaded, L1_Crude) %>% 
  filter(Date >= yearmonth("2010 Feb"), Date <= yearmonth("2020 March")) %>% 
  model(lm = TSLM(Unleaded ~ L1_Crude))

augment(fit_Unleaded_1) %>% 
  ggplot(aes(x = Date)) +
  geom_line(aes( y = Unleaded, color = "Data")) +
  geom_line(aes(y = .fitted, color = "Fitted")) +
  labs(y = NULL,
       title = "Average price of Unleaded in Dollars per Gallon") +
  scale_color_manual(values = c(Data="black", Fitted="#D55E00")) +
  guides(color = guide_legend(title = NULL))
```

The residuals show an increasing trend and the ACF shows a significant lag at period 12. The histogram is left skewed. To remedy this I tried using more predictors. Adding more predictors from the other time series I downloaded seems to help make recent years around January 2020 a better fit. 
Next, I compared variety of models I thought would be applicable. The best model selected was TSLM(Unleaded ~ L1_Crude + L1_PDI + season(1 year)).

```{r}
augment(fit_Unleaded_1) %>%
gg_tsdisplay(.resid, plot_type="hist")
```


```{r, message = FALSE}
eia_months_lag %>% 
  select(Date, Unleaded, L1_Crude, L1_Unemp, L1_PDI, L1_CPI) %>% 
  filter(Date >= yearmonth("2010 Feb"), Date <= yearmonth("2020 March")) %>% 
  model(
    TSLM(Unleaded ~ L1_Crude + L1_Unemp + L1_PDI + L1_CPI)
    )%>% 
  augment() %>% 
  ggplot(aes(x = Date)) +
  geom_line(aes( y = Unleaded, color = "Data")) +
  geom_line(aes(y = .fitted, color = "Fitted")) +
  labs(y = NULL,
       title = "Average price of Unleaded in Dollars per Gallon") +
  scale_color_manual(values = c(Data="black", Fitted="#D55E00")) +
  guides(color = guide_legend(title = NULL))
```


```{r}
fit_eai_all <-  eia_months_lag %>% 
  select(Date, Unleaded, L1_Crude, L1_Unemp, L1_PDI, L1_CPI) %>% 
  filter(Date >= yearmonth("2010 Feb"), Date <= yearmonth("2020 March")) %>% 
  model(
    # Linear trend
    linear = TSLM(Unleaded ~ trend()),
    # Exponential trend
    exponential = TSLM(log(Unleaded) ~ trend()),
    # Piecewise linear trend
    piecewise = TSLM(Unleaded ~ trend(knots = c(2016, 2019))),
    trend_season =  TSLM(Unleaded ~ trend() + season()),
    Fit_simple_Crude = TSLM(Unleaded ~ L1_Crude), 
    Crude_and_Season = TSLM(Unleaded ~ L1_Crude + season()),
    fit_Crude_PDI_Lag1 =  TSLM(Unleaded ~ L1_Crude + L1_PDI),
    Crude_PDI_Season = TSLM(Unleaded ~ L1_Crude + L1_PDI + season(period = "1 year"))
    )
    
glance(fit_eai_all) %>% 
  select(.model, r_squared, adj_r_squared, AICc, CV) %>% 
  arrange(desc(adj_r_squared))
```

Here are the plots of the time series selected for the best model `Crude_PDI_Season`. To make the plots more readable,I only inlcuded dates from 2015 onwards. 

```{r}
# Time series used for best model TSLM(Unleaded ~ L1_Crude + L1_PDI + season(1 year))
eia_months_lag %>% 
  filter(Date >= yearmonth("2015 Jan"), Date <= yearmonth("2020 March")) %>%      
  autoplot(vars(Unleaded, L1_Crude, L1_PDI))
```


The equation of the best model is $\widehat{Unleaded} =  -0.018 + 02145L1Crude + 0.00014L1PDI + season$.
 
The time plot of the residuals shows that the variation of the residuals stays much the same across the  data and therefore the residual variance can be treated as constant. This can also be seen on the histogram of the residuals. The histogram suggests that the residuals are normal. Consequently, forecasts from this method will probably be quite good and prediction intervals that are computed assuming a normal distribution are accurate.
Looks like residuals are normal and random error, they appear to be white noise. The short term correlations are significant however.
```{r}
# fit_eai_all %>% select(Crude_PDI_Season) %>% report() 
fit_eai_all %>% select(Crude_PDI_Season) %>% 
  augment() %>%  gg_tsdisplay(.resid, "histogram") 
```

To make a forecast for 12 monthly periods I used data from April 2020 to March 2021 to assess the accuracy of the forecast. 
 L1_Unemp, L1_CPI
```{r}
# April 2020 to March 2021
eai_1year <- eia_months_lag %>% 
    select(Date, Unleaded, L1_Crude, L1_PDI) %>% 
    filter(Date > yearmonth("2020 March"))

fit_unleaded_best <- fit_eai_all %>% select(Crude_PDI_Season) 

fc_unleaded_best <- fit_unleaded_best %>%  forecast(eai_1year)

# Plot the forecasts
eia_months_lag %>% filter(Date >= yearmonth("Jan 2015")) %>% 
  autoplot(Unleaded) +
  autolayer(fc_unleaded_best, level = NULL) +
   labs(y = "$US/Gallon",
       title = "US Monthly Average Gas Forecasted prices for 12 months",
       subtitle = "March 2020 to March 2021") 
  
```

The prediction intervals are not too wide, but showing them on the plot made it difficult to see the actual time-series values for unleaded for comparison. 
```{r}
hilo(fc_unleaded_best) %>% unpack_hilo(c("80%", "95%")) %>% select(.model,Date, "80%_lower":"95%_upper")
```
```{r}
accuracy(fc_unleaded_best, eai_1year ,
  measures = list(
    point_accuracy_measures,
    interval_accuracy_measures,
    distribution_accuracy_measures
  )
) %>% select(.model:MAPE)
```


## 2 Exponential Smoothing

For exponential smoothing I wanted to determine any trend and seasonality components. 
Here I show the data again for US average unleaded gas price with a superimposed trend fitted using the `STL()` and LOESS method. 

```{r}
decomp_model <- eia_months_lag %>% select(Date, Unleaded) %>%
   filter(Date >= yearmonth("2010 Feb"), Date <= yearmonth("2020 March")) %>% 
  model(stl = STL(Unleaded)) %>% 
  components() %>% 
  select(-.model) 
 decomp_model %>% autoplot(Unleaded, color="gray") +
  geom_line(aes(y=trend), color = "#D55E00") +
  labs(y = "$ per Gallon",title = "Average Price of Unleaded Gas: US")
```

```{r}
eia_months_lag %>% select(Date, Unleaded) %>%
   filter(Date >= yearmonth("2010 Feb"), Date <= yearmonth("2020 March")) %>% 
  model(stl = STL(Unleaded)) %>% components() %>% 
  autoplot() + labs(title = "ETS components")
```


I used the `ETS()` function to select the model by minimizing the AICc. The model selected is an Additive Errors model. I also tried some other models I could think of but they were not helpful. 

```{r}
fit_Unleaded_2a <-  eia_months_lag %>% select(Date, Unleaded) %>%
   filter(Date >= yearmonth("2010 Feb"), Date <= yearmonth("2020 March")) %>%
  model(auto = ETS(Unleaded),
        additive = ETS(Unleaded ~ error("A") + trend("A") +
                                                season("A", period = 12)),
        multiplicative = ETS(Unleaded ~ error("M") + trend("A") +
                                                season("M", period = 12))
  )

fit_Unleaded_2a %>% pivot_longer(everything(), names_to = "Model name",
                       values_to = "Orders")
```



```{r}
glance(fit_Unleaded_2a) %>% arrange(AICc)
```

```{r}
fit_Unleaded_2a %>% select(auto) %>%  report()
```


```{r}
fit_Unleaded_2a %>% select(auto) %>% gg_tsresiduals()
```

Forecasts 12 months ahead, from April 2020 to March 2021 showing the past three years.
```{r}
# forecast 12 months ahead  
fc_2a <- fit_Unleaded_2a %>% select(auto) %>% forecast(h = "1 year") 

fc_2a %>% 
  autoplot( eia_months_lag %>%  filter(Date > yearmonth("2020 Jan"))) 
```

```{r}
accuracy(fc_2a, eai_1year)
```

```{r}
hilo(fc_2a) %>% unpack_hilo(c("80%", "95%")) %>% select(.model,Date, "80%_lower":"95%_upper")
```



## 3 ARIMA

From the ACF I could see the correlations are declining in a linear fashion with some scallop shape around 12 months. Since the data are not stationary I took the second difference and checked the differences look like white noise.

```{r}
eia_months_lag %>% select(Date, Unleaded) %>% 
  filter(Date > yearmonth("2010 Jan"),  Date <= yearmonth("2020 March")) %>% 
   gg_tsdisplay(Unleaded, plot_type = "partial")
```


```{r, message=FALSE, warning=FALSE}
eia_months_lag %>% select(Date, Unleaded) %>%
  filter(Date > yearmonth("2010 Jan"), Date <= yearmonth("2020 March")) %>% 
gg_tsdisplay(difference(Unleaded, differences = 1), plot_type="partial", lag = 36)
```


The first difference shows significant spikes at lag 4, 8, and 12 suggest a non-seasonal moving average, MA(4), component. The significant spike at lag 12 in the ACF suggests a seasonal MA(1) component. This gives us a starting point of an ARIMA(0,1,4)(0,1,1)[12] indicating a first difference, a seasonal difference, a non-seasonal MA(1) component and a seasonal MA(1) component. This guess from the ACF plots is close to the one selected automatically ARIMA(0,1,4)(1,0,0)[12].


```{r}
fit_3a <- eia_months_lag %>% select(Date, Unleaded) %>% 
  filter(Date > yearmonth("2010 Jan"), Date <= yearmonth("2020 March")) %>% 
  model(
    arima011011 = ARIMA(Unleaded ~ pdq(0,1,1) + PDQ(0,1,1)),
    auto = ARIMA(Unleaded, stepwise = FALSE, approx = FALSE)   # takes longer but more accurate
  )

fit_3a %>% pivot_longer(everything(), names_to = "Model name",
                       values_to = "Orders")

```


```{r}
glance(fit_3a) %>% arrange(AICc)
```

```{r}
fit_3a %>% select(.model = "auto" ) %>%  report()
```


The residuals for the best model are shown below. One small bust significant spike at lag 32 is still consistent with white noise. To be sure we use a Ljung-Box test, A large p-value confirms that the residuals are similar to white noise. 
```{r}
fit_3a %>% select(auto) %>% gg_tsresiduals(lag=36)
```



```{r}
augment(fit_3a) %>%  filter(.model == "auto") %>% 
  features(.innov, ljung_box, lag = 32, dof = 4)
```


This seasonal ARIMA model passes the required checks and is ready for forecasting. A Forecast from the model for the next 12 months are shown. 

```{r}
forecast(fit_3a, h = 12) %>% 
  filter(.model =='auto') %>% 
  autoplot( eia_months_lag %>%  filter(Date > yearmonth("2018 Jan"))) +
  labs(title = "US unleaded average gas price",
       y = "Dollars per Gallon")
```
The 80% prediction interval does capture the future prices so I think this is a good forecast. 



####  Neural network autoregression

The `NNETAR()` function fits a model with seasonal data, NNAR(2,1,2)$_{12}$ gives the 12 month lag for the first year, the first 2 lags of Unleaded and 2 hidden networks. 


```{r}
Neural_fit <- eia_months_lag %>% select(Date, Unleaded) %>%
   filter(Date >= yearmonth("2010 Feb"), Date <= yearmonth("2020 March")) %>% 
  model(NNETAR(Unleaded)) 

Neural_fit %>% report()
```


A 12 month forecast is plotted which is capturing the upward trend but with some dampening.
```{r}
fc_Neural <- Neural_fit %>%   forecast(h = 12) 
fc_Neural %>% 
  autoplot( eia_months_lag %>%  filter(Date >    yearmonth("2018 Jan")))
```


```{r}
accuracy(fc_Neural, eai_1year ,
  measures = list(
    point_accuracy_measures,
    interval_accuracy_measures,
    distribution_accuracy_measures
  )
)
```

